---
title: "Statistical Analysis of Designs with Repeated Measures by Linear Mixed Models"
author: "Livio Finos"
output:
  pdf_document:
    number_sections: yes
    toc: yes
  html_document:
    number_sections: yes
    toc_float: yes
    toc: yes
---

# Motivation
```{r setup, message=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Import

```{r}

load("./dataset/datiEEG_LMM_2x2.Rdata")
summary(dati)
```


- 10 Subjects,
- 2-levels factor `Chan`
- 2-levels factor `Condition`


## EDA

For  `Y`: P300 Average Amplitude

```{r}
library(ggplot2)
p <- ggplot(dati,aes(x=Chan,y=Y,fill=Condition))
p=p+geom_dotplot(binaxis = "y",position=position_dodge(0.8),stackdir = "center") +geom_boxplot(alpha=.1)
p
```

Is there a specificity of the Subject?

```{r}
dati$Chan_Condition=paste(sep = "_",dati$Chan,dati$Condition)

p <- ggplot(dati,aes(x=Chan_Condition,y=Y))
p+geom_point(aes(group = Subj, colour = Subj))+
  geom_line(aes(group = Subj, colour = Subj))+
   geom_boxplot(alpha=.1)
```

We subtract the Subject-specific effect (i.e. Subject's mean) to each observation.

```{r}
mod=lm(Y~Subj,data=dati)
# summary(mod)
Y=residuals(mod)
dati$Yc=as.vector(Y)



library(ggplot2)
p <- ggplot(dati,aes(Chan_Condition,Yc))
p+geom_point(aes(group = Subj, colour = Subj))+
  geom_line(aes(group = Subj, colour = Subj))+
   geom_boxplot(alpha=.1)
```

The dispersion of the data has been largely reduced.
This effect is the one taken in account by the models for repeated measures. 

# Mixed models

## IMPORTANT REMARK about contrasts in (mixed) linear models

```{r}
# VERY IMPORTANT:
contrasts(dati$Condition) <- contr.sum(2) #2 is the number of levels
contrasts(dati$Chan) <- contr.sum(2) #2 is the number of levels
```


## Intuition

```{r message=FALSE,echo=FALSE}
par(mfrow=c(1,3))
voidplot <- function(title){
  plot(0,0,col=0,xlim=c(0,1),ylim=c(0,1),xlab="X",ylab="Y",asp=1,main=title)
  abline(0,1,lwd=3)
}
voidplot("one for all")
voidplot("Subject-specific intercept")
intrc=rnorm(5)/3
intrc[5]=-sum(intrc[-5])
for(i in 1:5) abline(intrc[i],1,lwd=2,col=i+1,lty=2)

slopes=rnorm(5)/3
slopes[5]=-sum(slopes[-5])
voidplot("Subject-specific intercept + slope")
for(i in 1:5) abline(intrc[i],1+slopes[i],lwd=2,col=i+1,lty=2)
```

Mixed models allow to model Subject-specific (average) effect by assuming that it is randomly drawn from the dstribution of the population (which is normal).


I assume you are expert on mixed models, if not
<https://en.wikipedia.org/wiki/Mixed_model>   
and much more on:
<http://webcom.upmf-grenoble.fr/LIP/Perso/DMuller/M2R/R_et_Mixed/documents/Bates-book.pdf>  
and  
<https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf>


Due to the small size of the dataset, in our example we only explore the scenario with random intercept and fixed slope (i.e. a simpler model, less parameters).

## The model
Models with random effects can be defined as:
$$Y_{n\times 1} = X_{n\times p} B_{p\times 1} + Z_{n\times q} b_{q\times 1} + \varepsilon_{n\times 1}$$

where $$\varepsilon\sim\mathcal{N}( 0,\sigma^2 I_n)$$

In the models we will consider, the random effects are modeled
    as a multivariate normal random variable:
    $$b\sim\mathcal{N}(0,\Sigma_{q\times q}),$$

In a *linear mixed model* the Conditional distribution $(Y|\mathcal B=b)$ is    a *spherical* multivariate Gaussian.


In our case $n=\#Subj \times \#Chan \times \#Condition=10\times 2 \times 2 = 40$.
$X$ is the matrix of (dummified) predictors. 
$Z$ can take many dimensions and values. Examples follow.

**Random effect for Subject (Random Intercept)**

$Z$ is the matrix of dummy variables of the column `dati$Subj`.

```{r message=FALSE}
library(lmerTest)

mod=lmer(Y~ Condition*Chan +(1|Subj),data=dati)

summary(mod)

car::Anova(mod,type=3)
```



## Plotting tools

for the first model:
```{r message=FALSE}
library(effects)
plot(allEffects(mod))

#plot random effects:
require(lattice)
qqmath(ranef(mod, condVar=TRUE))
```



An alternative plotting tool:

```{r message=FALSE}
library(sjPlot)
library(ggplot2)
plot_model(mod, type = "pred", terms = c("Chan","Condition"))
```


## Validity of the assumptions


* Independence of the residuals?
* Normality of the residuals?
* Homoscedasticity of the residuals (i.e. same variance between Subject/Condition/Chan?
* outliers?

* Leaverage? (influential observations)

Please, do not test for normality, for homoscedasticity, sphericity etc. 

Use Exploratory data Analysis, instead!


```{r}
dati$residuals=residuals(mod)
p <- ggplot(dati, aes(x=Chan_Condition, y=residuals,fill=Subj)) +   geom_boxplot() 
p

p <- ggplot(dati, aes(x=Subj, y=residuals,col=Subj)) +   geom_boxplot()+  geom_point(aes(group = interaction(Subj)))
p
```

<!-- ## Contrasts and post-hoc -->

<!-- ### Post-hoc -->
<!-- ```{r} -->
<!-- library(multcomp) -->
<!-- summary(glht(mod, linfct = mcp(Chan = "Tukey"))) -->
<!-- ``` -->

<!-- ### Custom contrasts -->
<!-- An example:    -->

<!-- - neutral vs object in O1 (left) -->
<!-- - disgust vs neutral in O1 (left) -->
<!-- - fear vs neutral in O1 (left) -->
<!-- - happy vs neutral in O1 (left) -->

<!-- ```{r} -->
<!-- library(multcomp) -->
<!-- ncoeff=length(coefficients(mod2)[[1]]) -->
<!-- contr <- rbind("n - o" = c(0,0,0,0,0,0,1,-1, rep(0,ncoeff-8)), -->
<!--                "d - n" = c(0,0,0,0,0,0,-1, rep(0,ncoeff-7)), -->
<!--                  "f - n" = c(0,0,0,0,1,0,-1, rep(0,ncoeff-7)),  -->
<!--                  "h - n" = c(0,0,0,0,0,1,-1, rep(0,ncoeff-7))) -->
<!-- compa= glht(mod2, linfct = contr) -->
<!-- summary(compa, test = adjusted("none")) -->

<!-- # with multiple comparisons -->
<!-- summary(compa) -->
<!-- ``` -->


# (minimal) Bibliography

Jonathan Baron (2011) Notes on the use of R for psychology experiments and questionnaires
<https://www.sas.upenn.edu/~baron/from_cattell/rpsych/rpsych.html>

and Course materal of   
ST 732, Applied Longitudinal Data Analysis, NC State University
by Marie Davidian
<https://www.stat.ncsu.edu/people/davidian/courses/st732/notes/chap5.pdf>
from <https://www.stat.ncsu.edu/people/davidian/courses/st732/>


About Type I, II, III SS:    <https://mcfromnz.wordpress.com/2011/03/02/anova-type-iiiiii-ss-explained/>

About Mixed models:   
<http://webcom.upmf-grenoble.fr/LIP/Perso/DMuller/M2R/R_et_Mixed/documents/Bates-book.pdf>  
and  
<https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf>
