---
title: "Permutation Testing with covariates: a lab"
author: "Livio Finos"
date: "University of Padova"
output: 
  html_document:
    toc: yes
    # toc_float: yes
    number_sections: yes
  pdf_document:
    keep_tex: yes
    number_sections: yes
    toc: yes
  beamer_presentation: default  
  ioslides_presentation:
    logo: figs/logoUnipd.jpg
---

# Intro

The following examples refers to the materials exposed in 
<https://github.com/livioivil/inference_basics/blob/master/perm_covariates_22.pdf>

```{r}
n=6
#iid error: example of var/cov matrix:
Sigma=3*diag(n)
# \Sigma=\sigma^2 I


I=diag(n)
#A (not so) random permutation matrix
Pi=matrix(0,n,n)
for(i in seq(1,n,by=2))
  Pi[i,i+1]<-Pi[i+1,i]<-1

head(Pi)


```



# Coviarates: only the intercept

```{r}
Z=matrix(1,n,1)

H=Z%*%solve(t(Z)%*%Z)%*%t(Z)
IH=I-H


```

Under the null hypo: $f((I-H)y)=f(\Pi (I-H) y)$? YES!

The mean of each obs is 0 after residualization, the variance is $\Pi(I-H)\Sigma(I-H)Pi=\sigma^2 \Pi(I-H)\Pi$

```{r}
#\sigma^2 doesn't change over \Pi, compare the remaining part:

IH
Pi%*%IH%*%Pi
```
# (Continuos) Covariates

```{r}
(Z=cbind(1,rnorm(n)))
Z=model.matrix(~0+.,data=data.frame(Z))

Zcont=Z # stored for later

H=Z%*%solve(t(Z)%*%Z)%*%t(Z)
(IH=I-H)


```

Under the null hypo: $f((I-H)y)=f(\Pi (I-H) y)$? No, in general!

The mean of each obs is 0 after residualization, the variance is $\Pi(I-H)\Sigma(I-H)Pi=\sigma^2 \Pi(I-H)\Pi$

```{r}
IH
Pi%*%IH%*%Pi
```


# Discrete Covariates

```{r}
(Z=factor(rep(1:2,each=n/2)))
(Z=model.matrix(~0+.,data=data.frame(Z)))


H=Z%*%solve(t(Z)%*%Z)%*%t(Z)
(IH=I-H)


```

Under the null hypo: $f((I-H)y)=f(\Pi (I-H) y)$? No, in general!

The mean of each obs is 0 after residualization, the variance is $\Pi(I-H)\Sigma(I-H)Pi=\sigma^2 \Pi(I-H)\Pi$

```{r}
IH
Pi%*%IH%*%Pi
```

But we can restrict the permutations within the same levels of $Z$

```{r}
Pi3=I
Pi3[1:3,1:3]=Pi3[3:1,1:3]
Pi3[4:6,4:6]=Pi3[6:4,4:6]
Pi3
Pi3%*%IH%*%Pi3
```


# Paired sample modeled by dummy variables

Paired sample are a special case of categorical covariates:

```{r}
(Z=factor(rep(1:(n/2),each=2)))
(Z=model.matrix(~0+.,data=data.frame(Z)))


H=Z%*%solve(t(Z)%*%Z)%*%t(Z)
(IH=I-H)

```

Under the null hypo: $f((I-H)y)=f(\Pi (I-H) y)$ only if one restrict the permutations within the same levels of $Z$:

```{r}
#(Pi already permutes within subjects)
IH
Pi%*%IH%*%Pi
```

# Decomposing I-H

```{r}
Z=Zcont
H=Z%*%solve(t(Z)%*%Z)%*%t(Z)
(IH=I-H)

ei=eigen(IH)
ei$values
ei$vectors

```

